Association between trauma severity models and opportunities for improvement: A retrospective cohort study

### 2022-09-15 studyprogressreport.Rmd
final fixes for today i swear

Introduction
============
Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Nonfatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].

@@

The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review. In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal considering that it cannot be known whether a case was mismanaged prior to it being reviewed. This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo]. Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo]. The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours. The latter benefit is not insignificant considering the large caseload found in most urban trauma centres. The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines]. Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo]. Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi]. Another potential audit filter, that is already used at some clinics [@whoguidelines], are trauma severity scores.

@@

Trauma scoring has become a cornerstone of modern trauma QI programs [@historytss]. One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines]. TSS also allow clinics to compare their outcomes against an established benchmark [@whoguidelines]. However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from. Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality]. This weakness is one reason why models continue to be developed and improved on. Another reason is the efforts to establish which risk factors best predict mortality; different models use different physiological and anatomical parameters.

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide [@trissbetter]. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries [@trissbetter]. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score can be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].

>>>>

Introduction
============
Traumatic injury is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of physical injury has led to trauma being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma].

@@

The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review. In an ideal world, only cases that were actually mismanaged would be selected for review; this is, of course, an impossible ideal considering that it cannot be known whether a case was mismanaged prior to it being reviewed. This reliance on case selection is a weakness of M&M review and has led to considerable efforts to develop strategies to standardise case selection [@tqipanglo]. Despite this, the process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo]. The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours. The latter benefit is not insignificant considering the large caseload found in most urban trauma centres. The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines]. Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo; @developqi]. Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi]. Another potential audit filter, that is already used at some clinics [@whoguidelines], is trauma severity scoring.

@@

Trauma scoring has become a cornerstone of modern trauma QI programs [@historytss]. One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines]. TSS also allow clinics to compare their outcomes against an established benchmark [@whoguidelines]. However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from. Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality]. This weakness is one reason why models continue to be developed and improved on. Another reason is that efforts to establish which risk factors best predict mortality are still ongoing; different models use different physiological and anatomical parameters.

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide [@trissbetter]. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries [@trissbetter]. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].
